---
title: "dplyr_tutorial"
output: html_document
---

## sources 
- This tutorial is from  https://www.youtube.com/watch?v=jWjqLW-u3hc
- View the R Markdown document: http://rpubs.com/justmarkham/dplyr-tutorial
- Download the scource document: https://github.com/justmarkham/dplyr-tutorial

 
## dplyr functionality 
- five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarise` (and `group_by`)
- work with data sroted in databases and [data tables](http://datatable.r-forge.r-project.org/)
- joins: inner join, left join, semi-join, anti-join
- window functions for calculating ranking, offsets, and more
- [Better than plyr](http://blog.rstudio.org/2014/01/17/introducing-dplyr/) if working with data frames. 

## Loading dplyr and an example dataset  
- dplyr will mask a few base functions

```{r}
# load packages
.libPaths("/home/ych/R//x86_64-pc-linux-gnu-library/3.0/")
library(dplyr)
```

```{r}
suppressMessages(library(dplyr))
library(hflights)

## explore data 
data(hflights)
head(hflights)
```

- `tbl_df` creats a "local data frame"
- local data frame is simply a wrapper for a data frame that prints nicely

```{r}
# convert to local data frame
flights = tbl_df(hflights)

# printing only shows 10 rows and as many columns as can fit on your screen
flights
```

```{r}
# you can specify that you want to see more rows
print(flights, n = 20)

# convert to a normal data frame to see all of the columns
data.frame(head(flights))
```

## filter: keep rows matching criteria
- base R approach to filtering forces you to repeat the data frames's name
- dplyr approach is simpler to write and read 
- Command structure (to all dplyr verbs):
    - first argument is a data frame 
    - return value is a data frame 
    - nothing is modified in place 
- Note: dplyr generally does not preserve row names, so if use rownames, it is better to write it into data frame explitly.  

```{r}
# base R approach to view all flights on January 1
flights[flights$Month == 1 & flights$DayofMonth == 1, ]
```

```{r}
# use pipe for OR condition
filter(flights, UniqueCarrier == "A" | UniqueCarrier == "UA")
```


```{r}
# you can also use %in% operator
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```

## select: Pick columns by name
- Base R approach is awkward to type and to read
- dplyr approach uses similar syntax to `filter`
- Like a SELECT in SQL

```{r}
# base R approach to select DepTime, ArrTime, and FlightNum columns
flights[, c("DepTime", "ArrTime", "FlightNum")]
```

```{r}
# dplyr approach 
select(flights, DepTime, ArrTime, FlightNum)
```

```{r}
# use colon to select multiple contiguous columns, and use `contains` to match column by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name 
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))
```

## "Chaining" or "Pipelining"
- this is not verbs
- Usual way to perform multiple operations in one line by nesting 
- Can write commands in a natural order by using the `%>%` infix operator (which can be pronounced as "then")
```{r}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)
```

```{r}
# chaining method 
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  filter(DepDelay > 10)
```

- Chaining increasing readability significantly when there are many commands
- Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
- Can be used to replace nesting in R commands outside of dplyr 

```{r}
# create two vectors and calculate Euclidian distance between them 
x1 <- 1:5;
x2 <- 2:6
sqrt(sum((x1 - x2)^2))
```

```{r}
#Chaining method 
(x1 - x2)^2 %>% sum() %>% sqrt()
```

## arrange: Reorder rows
```{r}
# base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]
```

```{r}
# dplyr approach
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  arrange(DepDelay)
```

```{r}
# use `desc` for descending 
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  arrange(desc(DepDelay))
```

## mutate: Add new variables
- Creat new variables that are functions of existing variables 
```{r}
# base R approach to create a new variable Speed (in mph)
flights$Speed = flights$Distance / flights$AirTim*60
flights[, c("Distance", "AirTime", "Speed")]
```

```{r}
# dplyr approach (prints the new variable but does not store it)
flights %>%
  select(Distance, AirTime) %>%
  mutate(Speed = Distance/AirTime*60)
```

```{r}
# store the new variable 
flights <- flights %>% mutate(Speed = Distance/AirTime*60)
```

## summarise: Reduce variables
- Primarily useful with data that has been grouped by one or more variables
- `group_by` creats the groups that will be operated on 
- `summarise` uses the provided aggregation function to summarise each group

```{r}
# base R approaches to calculate the average arrival delay to each destination
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm = TRUE)))
head(aggregate(ArrDelay ~ Dest, flights, mean))
```

```{r}
# dplyr approach: creat a table grouped by Dest, and then summarise each grop by taking the mean of ArrDelay
flights %>%
  group_by(Dest) %>%
  summarise(avg_delay = mean(ArrDelay, na.rm = TRUE))
```

- `summarise_each` allows you to apply the same summary function to multiple columns at once
- Note: `mutate_each` is also available

```{r}
# for each carrier, calculate the percentage of flights cencelled or diverted 
flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(mean), Cancelled, Diverted)
```

```{r}
# for each carrier, calculate the minimum and maximum arrival and departure delays
flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(min(., na.rm = TRUE), max(., na.rm = TRUE)), matches("Delay"))
```

- Helper function `n()` counts the number of rows in a group
- Helper function `n_distinct(vector)` counts the number of unique items in that vector 
```{r}
# for each day of the year, count the total number of flights and sort in descending order 
flights %>%
  group_by(Month, DayofMonth) %>%
  summarise(flight_count = n()) %>%
  arrange(desc(flight_count))
```

```{r}
# rewrite more simply with the `tally` function
flights %>%
  group_by(Month, DayofMonth) %>%
  tally(sort = TRUE)
```

```{r}
# for each destination, count the total number of flights and the number of distinct planes that flew there 
## summarise - > n() ==> tally
flights %>%
  group_by(Dest) %>%
  summarise(flight_count = n(), plane_count = n_distinct(TailNum))
```

- Grouping can sometimes be useful without summarising 
```{r}
# for each destination, show the number of cancelled and not cancelled flights
flights %>%
  group_by(Dest) %>%
  select(Cancelled) %>%
  table() %>%
  head()
```

## Window Function 
- Aggregation function (like `mean`) takes n inputs and returns 1 value
- Window function takes n inputs and return n values 
    - Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`)
    
```{r}
# for each carrier, calculate which two days of the year they had their longest departure delays 
# note: smallest (not largest) value is ranked as 1. so you have to use `desc` to rank by largest value 
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  filter(min_rank(desc(DepDelay)) <= 2 ) %>%
  arrange(UniqueCarrier, desc(DepDelay))
```

```{r}
## rewrite more simply with the `top_n` function
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  top_n(2) %>%
  arrange(UniqueCarrier, desc(DepDelay))
```

```{r}
# for each month, calculate the number of flights and the change from the previous month
flights %>%
  group_by(Month) %>%
  summarise(flight_count = n()) %>%
  mutate(change = flight_count - lag(flight_count))
```

```{r}
# rewrite more simply with the `tally` function
flights %>%
  group_by(Month) %>%
  tally() %>%
  mutate(change = n - lag(n))
```

## Other Useful Convenience Functions
```{r}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)
```

```{r}
flights %>% sample_frac(0.25, replace = TRUE)
```

```{r}
# base R approach to view structure of an object
str(flights)
```

```{r}
glimpse(flights)
```

## Connecting to Databases
- dplyr can connect to a database as if the data was loaded into a data frame 
- Use the same ayntax for local data frames and databases
- Only generates SELECT statements
- Currently supports SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB
- Example below is bases upon an SQLite database containing the hflights data
    - Instructions for creating this database are in the databases vignette

## Going deeper with dplyr
## Loading dplyr and the nycflights 13 dataset
```{r}
rm(flights)
```
```{r}
# load packages
suppressMessages(library(dplyr))
library(nycflights13)
# print the flights dataset from nycflights13
flights
```
## Chossing columns: select,rename
```{r}
# besides just using select to pick columns
flights %>%
  select(carrier, flight)
```
```{r}
# use the minus sign to hide columns
flights %>%
  select(-month, -day)
```
```{r}
# hide a range of columns
flights %>%
  select(-(dep_time:arr_delay))
```
```{r}
# hide any column with a matching name
flights %>%
  select(-contains("time"))
```
```{r}
# pick columns using a character vector of column names
cols = c("carrier","flight","tailnum")
flights %>%
  select(one_of(cols))
```
```{r}
# select() can be used to rename columns, though all columns not mentioned are dropped
flights %>%
  select(tail = tailnum)
```
```{r}
# rename() does the same thing, except all columns not mentioned are kept
flights %>%
  rename(tail = tailnum)
```

## Choosing rows: filter, between, slice, sample_n, top_n, distinct
```{r}
# filter() supports the use of multiple conditions
flights %>%
  filter(dep_time >=600, dep_time <= 605)
```
```{r}
# between() is a concide alternative for determing, if numeric values fall in a range 
flights %>%
  filter(between(dep_time, 600, 605))

# side note: is.na() can also be useful when filtering
flights %>%
  filter(!is.na(dep_time))
```
```{r}
# slice() filters rows by position
flights %>%
  slice(1000:1005)
```

```{r}
# keep the first three rows within each group
flights %>%
  group_by(month, day) %>%
  slice(1:3)
```
```{r}
# sample three rows from each group: randomly
flights %>%
  group_by(month, day) %>%
  sample_n(3)
```

```{r}
# keep three rows from each group with the top dep_delay
flights %>%
  group_by(month, day) %>%
  top_n(3, dep_delay)
```
```{r}
# also sort by dep_delay within each group
flights %>%
  group_by(month, day) %>%
  top_n(3, dep_delay) %>%
  arrange(desc(dep_delay))
```

```{r}
# unique rows can be identified using unique() from base
flights %>%
  select(origin, dest) %>%
           unique()
```

## Adding new variables: mutate, transmute, add_rownames
```{r}
# mutate() creates a new variable (and keep all existing variables)
flights %>%
  mutate(speed = distance/air_time*60)
```
```{r}
# transmute() only keeps the new variable
flights %>%
  transmute(speed = distance/air_time * 60)
```
```{r}
# example data from with row names
mtcars %>%
  head()
```
```{r}
# add_rownames() turns row names into an explicit variable
mtcars %>%
  add_rownames("model") %>%
  head()
```
```{r}
# side note: dplyr no longer prints row names(ever) for local data frames
 mtcars %>%
  tbl_df()
```
## Grouping and counting: summarise, tally, count, group_size, n_groups, ungroup
```{r}
# summarise() can be used to count the number of rows in each group
flights %>%
  group_by(month) %>%
  summarise(cnt = n())
```
```{r}
# tally() and count() can do this more conscisely
flights %>%
  group_by(month) %>%
  tally()

flights %>%
  count(month)
```

```{r}
# can sort by the count
flights %>%
  group_by(month) %>%
  summarise(cnt = n()) %>%
  arrange(desc(cnt))
```
```{r}
# tally() and count() have a sort parameter for this purpose
flights %>% group_by(month) %>% tally(sort=TRUE)
flights %>% count(month, sort=TRUE)
```
```{r}
# sum over a specific variable instead of simply counting rows
flights %>%
  group_by(month) %>%
  summarise(dist = sum(distance))
```
```{r}
# tally() and count() have a wt parameter for this purpose 
flights %>%
  group_by(month) %>%
  tally(wt = distance)

flights %>%
  count(month, wt = distance)
```

```{r}
# group_size() returns the counts as a vector
flights %>%
  group_by(month) %>%
  group_size()
```
```{r}
# n_groups() simple reports the number of groups
flights %>%
  group_by(month) %>%
  n_groups()
```
```{r}
# group_ by two variables, summarise, arrange(output is possibly confuding)
flights %>%
  group_by(month, day) %>%
  summarise(cnt = n()) %>%
  arrange(desc(cnt)) %>%
  print(n = 40)
```

```{r}
# ungroup() before arranging to arrange across all groups
flights %>%
  group_by(month, day) %>%
  summarise(cnt = n()) %>%
  ungroup() %>%
  arrange(desc(cnt))
```

## Creating data frames: data_frame
`data_frame(0` is a better way than `data.frame()` for creating data frames. 
- use previously defined columns to compte new columns
- never ceerces column types
- new munges columns names
- never adds row names
- only recycles length 1 input
- returns a local data frame ( a tbl_df)

```{r}
# data_frame() example
data_frame(a = 1:6, b = a*2, c = 'string', 'd+e' = 1) %>% glimpse()
```

```{r}
# data.frame() exaple 
data.frame(a = 1:6, c = 'string', 'd+e' = 1) %>% glimpse()
```

## Joiing(merging) tables: left_join, right_join, inner_join,ffull_join, semi_join, anti_join
```{r}
# create two simple data frames
(a = data_frame(color = c("green","yellow","red"), num = 1:3))
```
```{r}
(b <- data_frame(color = c("green","yellow","pink"), size = c("S","M","L")))
```
```{r}
# only include observations found in both "a" and "b" automatically joins on variables that appera in both tables
inner_join(a,b)
```
```{r}
# include observations found in either "a" or "b"
full_join(a,b)
```

```{r}
# include all observations found in "a"
left_join(a,b)
```

```{r}
# include all observations found in "b"
right_join(a,b)
```
```{r}
# right_join(a,b) is identical to left_join(b,a) except for column ordering
left_join(b,a)
```
```{r}
# filter "a" to show observations that match "b"
semi_join(a,b)
```
```{r}
# filter "a" to only show obs that don't match "b"
anti_join(a,b)
```

```{r}
# sometimes matching variables don't have identical names
b = b%>% rename(col = color)
# specify that the join should occur by matching "color" in "a" with "col" in "b"
inner_join(a,b,by = c("color" = "col"))
```

## Viewing more output: print, View
```{r}
# specify that you want to see more rows
flights %>%
  print(n  = 15)
```
```{r}
# specify that you want to see ALL rows (don't run this!)
flights %>% print(n = Inf)
```
```{r}
# specify that you want to see all columns
flights %>% print(width = Inf)
```
```{r}
# show up to 1000 rows and all columns
flights %>% View()

# set option to see all columns and fewer rows
options(dplyr.width = Inf, dplyr.print_min = 6)

# reset options (or just close R)
options(dplyr.width = NULL, dplyr.print_min = 10)
```
